EAST: {\underline E}xpression Fragment {\underline A}ssembly from {\underline S}panning {\underline T}rees

### 2011-06-08 peace/papers/EAST/east.tex
Major modifications to EAST manuscript.

\NeedsTeXFormat{LaTeX2e}[1995/12/01]
\documentclass[10pt]{bmc_article}

@@

\author{Yuan Zhang$^1$
  \email{Yuan Zhang - zhangy9@muohio.edu}
  \and
  Dhananjai M Rao$^1$
  \email{Dhananjai Rao - raodm@muohio.edu}
  \and
  Mufit Ozden$^1$
  \email{Mufit Ozden - ozdenm@muohio.edu}
  \and
  James C Moler$^1$
  \email{James Moler - molerjc@muohio.edu}
  \and
  Chun Liang\correspondingauthor$^{1,2}$
  \email{Chun Liang\correspondingauthor - liangc@muohio.edu}
  and
  John E. Karro\correspondingauthor$^{1,3}$
  \email{John E. Karro\correspondingauthor - karroje@muohio.edu}
}

@@

\address{
    \iid(1)Department of Computer Science and Software Engineering, Miami University, Oxford OHIO, USA \\
    \iid(2)Department of Botanty, Miami University, Oxford OHIO, USA \\
    \iid(3)Department of Microbiology, Miami University, Oxford OHIO, USA
}

@@

\begin{abstract}
        % Do not use inserted blank lines (ie \\) until main body of text.
\end{abstract}

@@

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Introduction}
The {\it de novo} assembly of RNA transcripts is an important and
computationally challenging problem, made more difficult by the advent
of Next Generation Sequencing technologies such as 454 or Illumina
\cite{Nagaraj07,Rao10}.  Given the completion of a large-scale sequencing run of
multiple mature transcripts of a given organism, the investigator is
left with a large set of sequence fragments, but no ordering or even
transcript association information; these fragments must be clustered
by transcript then assembled before the sequencing output can be truly
useful.  If the genome has not been sequenced, this must be done {\it
  de novo} -- using only the sequence information extrapolated from
the sequences themselves.  Even when the genomic sequence is known,
there are a number of advantages to ignoring the known sequence
information and starting from scratch.  Using an existing sequence as
a template for sequencing in population studies has the potential to
introduce bias.  And when looking specifically for genes, a known
sequence may be of limited use: the target transcripts may be
undiscovered, and the splicing of introns before sequencing can
complicate the approach of mapping fragments to the genomic sequence.
In short, there is significant value in tools that can reassemble
fragment data into the original assembly without requiring access to
genomic information.

The problem of {\it de novo} transcript assembly has been addressed a
number of times in the literature, with initial approaches
concentrating on the longer reads produced by traditional Sanger
Sequencing techniques, and newer approaches looking at produces of
Next Generation Sequencing (NGS) technologies.  A list of the tools
most successfully addressing these problems include \capthree, \tgicl,
\velvet, and \mira, each excelling within their own context
\cite{Huang99,Pertea03,Chevreux04,Zerbino08}.  \capthree\/ and
\tgicl\/ use alignment-based strategies that produce high quality
results when applied to the output of Sanger sequencing technologies.
\velvet, based on the modeling of the problem with a {\it de bruijn}
graphs, was designed to handle short read sequences.  \mira\/ is a
multi-pass DNA assembler based on an overlap-layout-consensus
strategy.  [STILL NEED TO SAY SOMETHING ABOUT MIRA AND HYBRID SETS.]

In this paper we present \east, a tool for the assembly of both
Sanger, NGS and hybrid data with the capacity to incorporate quality
scores when available.  \east\/ is based on a novel minimum spanning
tree (MST) approach to compute assembly.  The tool works in tandem with
the \peace\/ clustering tool \cite{Rao10}: \peace\/ clusters
fragments by transcript and produces the initial MSTs that will serve
as the basis for the \east\/ algorithm.  In the following we present a
comparison of \east\/ against those of \capthree, \tgicl, \velvet, and
\mira on both Sanger and NGS sequences, showing that \peast\/
generates the best quality results in the fastest runtime for Sanger
sequences, and that while certain tools out-preform \peast\/ under specific
conditions, it is the only tool able to produce constant quality results
for Sanger, 454, Illumina, {\it and} hybrid data sets.

@@

In assessing result quality, we must view the \peace\/ clustering tool
\cite{Rao10} and \east\/ as a combined process.  While most assembly
tools integrate these two jobs, \east\/ assumes clustering has been
preformed and requires the \peace\/ MST output.  While other cluster
tools could be substituted in to preform the actual clustering
(e.g. \wcd or \easycluster \cite{Hazelhurst08,Picardi09}), only
\peace\/ provides the requires MST output.  Hence for purposes of the
following section, we consider a single tool \peast, and assess the
quality of these results.

In order to assess the quality of the (\peace\/ +) \east\/ results, we compared
against four prominent assembly tools: \capthree, \tgicl,
\mira, and \velvet\/ \cite{Huang99,Pertea03,Chevreux04,Zerbino08}.  Tools were tested on data from three
sequencing technologies (Sanger, 454 and Illumina), using both
simulated output obtained from the \estsim\/ and \metasim\/ tools
\cite{Hazelhurst03,Richter08} and benchmark sets obtained from GenBank \cite{Benson10}.

\subsection*{Measuring Quality}
Our primary measure of quality was the {\bf normalized a-score}, based
on the {\it assembly score} (a-score) used in {\it Liang et al.}
\cite{Liang00}.  For an assembly of some original sequence, the
a-score is defined as:
$$\mbox{($2\times$sequence length) - ($5\times$number of
mismatches) - ($15 \times$number of indels)}$$
with a ``perfect'' score equal to twice the sequence length.  Our
normalized a-score modifies this definition as:
$$\frac{\mbox{($2\times$assembly length) - ($5\times$number of mismatches)
    - ($15\times$number of indels)}}{\mbox{($2\times$sequence
    length)}}$$ with a perfect score equal to 1.  This normalization
allows for the comparison of assembly quality between sequences of
different lengths. Further, using this score we can more accurately
quantify the quality of a partial reconstruction: if we are able to
reconstruct only a portion of a fragment, the normalized a-score will
take full consideration of the quality of the reconstruction, but then
bound the score by the relative size of the reconstruction.
(e.g. Perfectly reconstruction one fourth of the original sequence
will lead to an a score of $0.25$.)  In addition to a-score we also
look at: the number of contigs produced by a tool (which, in the
perfect solution, will be exactly equal to the number of transcripts
represented by the data set), normalized by the number of transcripts;
the number of singleton ESTs in the solution (i.e. ESTs unassigned to
any contig and thus, most cases, effectively thrown away), and the runtime of the tool.

\subsection*{Simulation Results}
Using simulated fragment sets allows us to both compare assemblies to
a known correct solution and control parameters in their model
(e.g. base-call error rate) to test the effect of variation.  All sets
were generated from a set of 100 zebra fish transcripts obtained from
{\it Hazelhurst et al.}  \cite{Hazelhurst08}.  For simulated Sanger
sequences we used the \estsim\/ tool, using its default model of EST
generation when not otherwise specified.  For simulated
454 and Illumina sets we used the \metasim\/ tool \cite{Richter08}, again using their
default model to simulate the errors associated with those
technologies.

\noindent {\bf Sanger Sequencing:} We first compare the assembly
quality of the tools on Sanger Sequencing, using \east, \capthree, and
\tgicl\/ (the tools designed to work with Sanger sequences), looking
at the a-score as a function of error rate for a range of coverage
depths.  For each parameter combination we applied each tool to 30
independently randomly generated simulated sets, derived from 15 zebra
fish genes, reporting the average normalized a-score.  In
Figure~\ref{sangerAscore} we look at normalized a-score as a function
of error rate, fixing coverage at 50, 30 and 10 and varying base-read
error rate from 0\% to 6\% for each of the three tools designed for
Sanger Sequencing.  While all three tools preform well when applied to
error free (i.e. impossible) sequencing technology, we find \east\/ to
be considerably more robust to base-call errors than the other tools.
Taking the mid-range coverage value of 30 as an example, at a 1\%
error rate we see \east, \capthree, and \tgicl\/ all achieving an
essentially perfect normalize a-score .  At a 3\% error rate \east\/
maintains its almost perfect normalized a-score, showing a 2\%
improvement over \tgicl\/ and a 21\% improvement over \capthree.  At a
4\% error rate \east\/ still maintains its near-perfect score (showing
a 54\% and 85\% improvement over \capthree\/ and \tgicl\/
respectively), and even at a 6\% error rate is still achieving an
normalized a-score of 0.95 (a 110\% improvement over the next closest
tool, \tgicl\/), meaning that at most 5\% of the bases were left out of
their contigs or were incorrectly reconstructed.

\begin{figure}[htb]
\centerline{\includegraphics[width=6in]{pics.d/ascore_sanger.pdf}}
\caption{Average normalized a-score as a function of base error rate for
  simulated Sanger sequencing.  \east=blue, \capthree=Black,
  \tgicl=green, \velvet=red, \mira=purple.}
\label{sangerAscore}
\end{figure}

Another useful metric is the number of contigs and number of singleton
fragments in the final result.  In Figure~\ref{contigsSanger} is a
plot of the average number of contigs per gene sequence (left column), and
the number of singletons (right column).  For the first we observe that all tools have
comparable results for lower error rates, but only \east\/ is able to
maintain the low score with increasing error rates -- generating an
perfect score in almost every case.  In minimizing the
number of singletons, we find that \east\/ does a better job than
\capthree\/ for error rates larger than $2\%$, while \tgicl does a
better job that \east\/ for error rates larger than $4\%$ to $5\%$.

\begin{figure}[htb]
\begin{minipage}{3in}
\includegraphics[width=3in]{pics.d/numContigs_sanger.pdf}
\end{minipage}
\begin{minipage}{3in}
\includegraphics[width=3in]{pics.d/numSingle_sanger.pdf}
\end{minipage}
\caption{Number of contigs (normalized to the number of transcripts)
  and number of singletons as a function of base call error
  rate for simulated Sanger sequencing.  \east=blue, \capthree=Black,
  \tgicl=green, \velvet=red, \mira=purple.  (Number of singles for
  \velvet\/ and \mira\/ not shown.)}
\label{contigsSanger}
\end{figure}

\noindent {\bf Next Generation Sequences:} In
Figure~\ref{nextgenAscore} we compare the normalized a-score for
simulated 454 and Illumina outputs (with sequence lengths averaging
250 and 62 bp respectively).  For these tests we varied coverage
levels but otherwise took the default Metasim model.  On 454 data we
see significant improvement in results quality in \east\/ for all
levels of coverage.  On Illumina data we see that \east\/ is
competitive with \velvet\/ at higher coverages, though is unable to
match that tool when coverage is sparse.

\begin{figure}[htb]
\centerline{\includegraphics[width=6in]{pics.d/ascore_nxtgen.pdf}}
\caption{Comparative normalized a-scores for 454 and Illumina sequence
  output, using the \metasim default model at different levels
  of coverage.  \east=blue, \capthree=Black, \tgicl=green,
  \velvet=red, \mira=purple.  }
\label{nextgenAscore}
\end{figure}

\noindent {\bf Hybrid:} In Figure~\ref{hybridAscore} we look at the
quality of results derived from applying each tool to a hybrid data
set of Sanger (40\%), 454 (40\%) and Illumina (20\%) sequences, again
showing normalized a-score as a function of base call error rate in
the Sanger portion.  Here we observe \east matching or outperforming
\capthree\/ and \mira\/ at moderate to high coverages.  At a coverage
level of 50, \east\/ maintains a normalized a-score hovering around
0.85 at all error levels (showing a 54\% improvement over \mira\/ at a
3\% base-call error rate).  \east\/ actually improves its performance
at a coverage level of 30 (hovering around a normalized a-score of
0.95, showing a 142\% improvement of Mira at the 3\% base-call error
rate), an oddity we attribute to the decreased absolute number of NGS
sequences with the smaller coverage.  At a coverage level of 10
\east's normalized a-score is around 0.37 -- doing considerably worse
that \mira\/ at low error rates (e.g. 0.94 v. 0.40 for error-free
bases), but showing a 52\% improvement over \mira\/ at the 3\% error
rate mark and in general showing much more robustness to error.
Remarkably, it is only in this situation that \capthree\/ is
competitive, with that tool achieving a normalized a-score of 0.62
(v. \east's 0.37) for the 3\% base call error rate, but loosing out to
\east at greater than a 4.5\% base call error rate -- a result that we
attribute to \east's problems with the 62 bp Illumina sequences (see below).

\begin{figure}[htb]
\centerline{\includegraphics[width=6in]{pics.d/ascore_hybrid.pdf}}
\caption{Comparative normalized a-scores a hybrid set of Sanger, 454
  and Illumina sequence output, varying the base call error rate for
  Sanger sequences and using the MetaSim default model for the 454 and
  Illumina sequences.  Each data set tested was composed of 40\%
  Sanger Sequences, 40\% 454 Sequences, and 20\% Illumina sequences.
  \east=blue, \capthree=Black, \tgicl=green,
  \velvet=red,\mira=purple.}
\label{hybridAscore}
\end{figure}

\noindent {\bf Tool Runtime:} As \east\/ assembly requires the
\peace\/ cluster (as opposed to the competing tools that do both
jobs), for \east\/ runtime we actually include the \peace\/
runtime. In Figures~\ref{runtime.fixed} we see the average runtime for
each of the experiments reflected in Figures
Figures~\ref{sangerAscore}-\ref{hybridAscore} for a coverage of 30,
while in Figure~\ref{runtime.coverage} we look at the average runtime
for each tool when applied to simulated data sets generated from a
single zebra fish gene at varying levels of coverage -- effectively
plotting the runtime as a function of fragment set size.  For Sanger
sequences, we see marked improvement of (\peace+)\east\/ over all
other tools for almost all coverage levels: for the smallest set
(coverage = 10) \east\/ has 1.9 speedup over \capthree\/ and a 3.9
speedup over \tgicl, while at a more moderate size (coverage=50) we
see speedups of 4.7 and 8.4.  This returns back down to 3.0 and 5.2 at the
largest set tested (coverage = 100), as can be seen in the bump around
a coverage of 85 in Figure~\ref{runtime.coverage}.  We are unable to
superficially explain this bump, but have noticed the same effect in
data sets derived from other genes as well.

\east\/ is less competitive on shorter sequence (an effect of the
$d^2$ bias of our algorithm, as discussed below), reducing its runtime
improvement for the 250 bp 454 sequences, severely reducing it for the
62 bp Illumina sequences, and also affecting the hybrid data sets (also
due to the presence of the Illumina sequence).  We argue that the high
quality of our results still makes \east\/ an appealing option for 454
and hybrid data (while acknowledging that cannot complete with
\velvet\/ on pure Illumina data), and also observe that while it has
not been implemented, the basic \east\. algorithm should easily admit
paralilization -- thus increasing the practical limit on data set size.

\begin{figure}[htb]
\centerline{\includegraphics[width=6in]{pics.d/runtime_fixedsize_sanger.pdf}}
\caption{Average runtime per trial for when generating data for
  Figures~\ref{sangerAscore}-\ref{hybridAscore}  when coverage = 30).
  Note that the exceptionally high speed of \velvet\/ on Illumina sequences make the
  red (center) bars in that plot essentially indistinguishable.
  (\peace+)\east=blue, \capthree=Black, \tgicl=green,
  \velvet=red,\mira=purple.}
\label{runtime.fixed}
\end{figure}

\begin{figure}[htb]
\centerline{\includegraphics[width=6in]{pics.d/runtime_bycoverage_all.pdf}}
\caption{The average runtime when applying each tool to a simulated
  data set derived from a specific zebrafish transcript of length
  14,625 bp.  As we decrease coverage the number of fragments drops,
  hence illustrating the relationship between tool runtime and
  fragment set size.  Each data point represents the average runtime
  from the application of the tool to at least 30 randomly generated
  simulated sets coming from the transcript at the given level of coverage.
  (\peace + )\east=blue, \capthree=Black, \tgicl=green,
  \velvet=red,\mira=purple.}
\label{runtime.coverage}
\end{figure}

@@

In this manuscript we have described the \east\/ algorithm and run a
comparison of the \peast combination against the best preforming tools
in the literature.  When compared against the fine benchmark tools on
normalized a-score and runtime metrics, the \peast\/ combination of
tools prove to be both the best tool for Sanger sequencing, the
best tool overall, and by far the most robust against sequencing
error.  While certain tools may do better than it in specific circumstances,
there is no tool that does better than it on multiple sequencing
technology platforms.  While its speed is slower when applied to short
sequence fragments, the quality of its results more then compensate.

>>>>

\NeedsTeXFormat{LaTeX2e}[1995/12/01]
\documentclass[10pt]{bmc_article}    
\usepackage[usenames]{color}

@@

\author{Yuan Zhang$^1$
  \email{Yuan Zhang - zhangy9@muohio.edu}
  \and
  Dhananjai M Rao$^1$
  \email{Dhananjai Rao - raodm@muohio.edu}
  \and
  Mufit Ozden$^1$
  \email{Mufit Ozden - ozdenm@muohio.edu}
  \and
  Jens Mueller$^2$
  \email{Jens Mueller - muellej@muohio.edu}
  \and
  Chun Liang\correspondingauthor$^{1,3}$
  \email{Chun Liang\correspondingauthor - liangc@muohio.edu}
  and
  John E. Karro\correspondingauthor$^{1,4.5}$
  \email{John E. Karro\correspondingauthor - karroje@muohio.edu}
}

@@

\address{
    \iid(1)Department of Computer Science and Software Engineering,  \\
    \iid(2)Informational Technology Services
    \iid(3)Department of Botanty, \\
    \iid(4)Department of Microbiology, \\
    \iid(5)Department of Statistics, Miami University, Oxford OHIO, USA
}

@@

\begin{abstract}
\end{abstract}

@@

%%%%%%%%%%%%%%%%
%% Background %%
%%
  \section*{Introduction} {\color{blue} The {\it de novo} assembly of
    RNA transcripts is an important and computationally challenging
    problem, made more difficult by the advent of Next Generation
    Sequencing technologies such as 454 and Illumina
    \cite{Nagaraj07,Rao10}.  Upon completion of
    transcriptome-sequencing, the investigator is left with a large
    set of sequence fragments but no ordering or transcript
    association information.  These fragments must be clustered by
    transcript and assembled before the sequencing output can be truly
    useful.  {\it de novo} assembly tools assemble the
    transcript out of the fragments without using genmic sequence
    information.  {\it de novo} tools are obivously necessary when working with a genome that is
    unsequenced or poorly annotated, but is also useful when there is
    limited knowledge of gene splice sites, as well as for the discovery
    of novel events \cite{Birol09,Robertson10}.}

\vspace{3mm}

The problem of {\it de novo} transcript assembly has been addressed a
number of times in the literature, with initial approaches
concentrating on the longer reads produced by traditional Sanger
Sequencing techniques, and newer approaches looking at produces of
Next Generation Sequencing (NGS) technologies.  A list of popular
tools addressing these problems include \capthree, \tgicl, \velvet,
and \mira, each excelling within their own context
\cite{Huang99,Pertea03,Chevreux04,Zerbino08}.  \capthree\/ and
\tgicl\/ use alignment-based strategies that produce high quality
results when applied to the output of Sanger sequencing technologies.
\velvet, based on the modeling of the problem with {\it de bruijn}
graphs, was designed to handle short read sequences.  \mira\/ is a
multi-pass DNA assembler based on an overlap-layout-consensus strategy
that also has an increased ability to assemble {\it hybrid sets}: sets
formed by the mixture of sequencing results from multiple technologies
\cite{MiraWeb}.

\vspace{3mm}

In this paper we present \east: a tool for the assembly of
Sanger, NGS and hybrid data with the capacity to incorporate quality
scores when available.  \east\/ is based on a novel Minimum Spanning
Tree (MST) approach to compute assembly.  The tool works in tandem
with the \peace\/ clustering tool \cite{Rao10}: \peace\/ clusters
fragments by transcript and produces the initial MSTs that will serve
as the basis for the \east\/ algorithm.  In this paper we present a
comparison of \east\/ against \capthree, \tgicl, \velvet, and
\mira\/ on both Sanger and NGS sequences, showing that \peast\/
generates the best quality results in the fastest runtime for Sanger
sequences.  Although certain tools out-perform \peace\/ under
specific conditions, we show that it is the only tool able to produce consistently
higher quality results for Sanger, 454, Illumina, {\it and} hybrid
data sets.  In this paper we discuss the results and relevant
computational methods.  \east\/ has been incorporated into the
\peace\/ clustering tools \cite{Rao10}; it can be installed
and run through a Java-based PEACE GUI available at
www.peace-tools.org.

@@

\textcolor{blue}{In assessing result quality, we must view the \peace\/ clustering tool
\cite{Rao10} and \east\/ as a combined process.  While most assembly
tools integrate these two jobs, \east\/ assumes clustering has been
performed and requires the \peace\/ MST output.  Hence for purposes of
the following section, we consider them as a single tool \peast, and
assess the quality of these results.}

\vspace{3mm}

In order to assess the quality of (\peace\/ +) \east\/ results, we compared
against four prominent assembly tools: \capthree, \tgicl,
\mira, and \velvet\/ \cite{Huang99,Pertea03,Chevreux04,Zerbino08}.  The tools were tested on data from three
sequencing technologies (Sanger, 454 and Illumina), using both
simulated output obtained from the \estsim\/ and \metasim\/ tools
\cite{Hazelhurst03,Richter08}.

\subsection*{Measuring Quality}
Our primary measure of quality was the {\bf normalized a-score}, based
on the {\it assembly score} (a-score) used in {\it Liang et al.}
\cite{Liang00}.  For an assembly of some original sequence, the
a-score is defined as:
$$\mbox{($2\times$sequence length) - ($5\times$number of
mismatches) - ($15 \times$number of indels)}$$
with a ``perfect'' score equal to twice the sequence length.  Our
normalized a-score modifies this definition as:
$$\frac{\mbox{($2\times$assembly length) - ($5\times$number of mismatches)
    - ($15\times$number of indels)}}{\mbox{($2\times$sequence
    length)}}$$ with a perfect score equal to 1.  This normalization
allows for the comparison of assembly quality between sequences of
different lengths. Further, using this score we can more accurately
quantify the quality of a partial reconstruction: if we are able to
reconstruct only a portion of a fragment, the normalized a-score will
take full consideration of the quality of the reconstruction, but then
bound the score by the relative size of the reconstruction.
\textcolor{blue}{(For example, were a tool only able to reconstruct
  one fourth of a transcript -- but able to reconstruct that fraction
  perfectly -- the resulting contig's a-score would be $0.25$.)}  In
addition we look at: the number of contigs produced by a tool
normalized by the number of transcripts (a value equal to 1 in a
perfect reconstruction); the number of singleton ESTs in the solution
(i.e. ESTs unassigned to any contig and thus, most cases, effectively
thrown away), and the runtime of the tool.

\subsection*{Simulation Results}
Using simulated fragment sets allows us to compare assemblies to
a known correct solution and control parameters in their model
(e.g. base-call error rate) to test the effect of variation.  All sets
were generated from a collection of 100 zebra fish transcripts
obtained from {\it Hazelhurst et al.}  \cite{Hazelhurst08}.  For
simulated Sanger sequences we used the \estsim\/ tool, using its
default model of EST generation when not otherwise specified
\cite{Hazelhurst03}.  For simulated 454 and Illumina sets we used the
\metasim\/ tool, again using their default model to simulate the
errors associated with those technologies \cite{Richter08}. {\bf
  [COMMENT: Will try to add a discussion of, or reference to, the
  default values here.]}

\vspace{3mm}

\noindent {\bf Sanger Sequencing:} \textcolor{blue}{We first compare the
  assembly quality of the tools on a Sanger Sequencing model, using
  \east, \capthree, and \tgicl\/, the tools designed to work with
  Sanger sequence.  The quality of any assembly will be directly
  affected by the {\it covreage} of the input: the average number of
  ESTs that cover a based.  Thus we investigate the a-score as a
  function of error rate at different coverage levels.}  For each
parameter combination we applied each tool to 30 independent, randomly
generated simulated sets, derived from 15 randomly chosen members of
our zebra fish gene set, reporting the average normalized a-score.  In
Figure~\ref{sangerAscore} we look at normalized a-score as a function
of error rate, fixing coverage at 50, 30 and 10 and varying base read
error rate from 0\% to 6\% for each of the three tools designed for
Sanger Sequencing.  While all three tools perform well when applied to
error free (i.e. impossible) sequencing technology, we find \east\/ to
be considerably more robust to base-call errors than the other tools.
Taking the mid-range coverage value of 30 as an example, at a 1\%
error rate we see \east, \capthree, and \tgicl\/ all achieving an
essentially perfect normalized a-score.  At a 3\% error rate \east\/
maintains its almost perfect normalized a-score, showing a 2\%
improvement over \tgicl\/ and a 21\% improvement over \capthree.  At a
4\% error rate \east\/ still maintains its near-perfect score (showing
a 54\% and 85\% improvement over \capthree\/ and \tgicl\/
respectively).  Even at a 6\% error rate \east\/ is still achieving a
normalized a-score of 0.95 (a 110\% improvement over the next closest
tool, \tgicl\/), meaning that at most 5\% of the bases were left out
of their contigs or were incorrectly reconstructed.

\begin{figure}[htb]
\centerline{\includegraphics[width=6in]{pics.d/ascore_sanger_norm.pdf}}
\caption{Average normalized a-score as a function of base error rate for
  simulated Sanger sequencing.  \east=blue, \capthree=Black,
  \tgicl=green. COMMENT: Need to change figure heading to Normalized a-score.}
\label{sangerAscore}
\end{figure}

In Figure~\ref{contigsSanger} is a
plot of the average number of contigs per gene sequence (left column), and
the number of singletons (right column).  For the first we observe that all tools have
comparable results for lower error rates, but only \east\/ is able to
maintain the low score with increasing error rates -- generating a
perfect score in almost every case.  In minimizing the
number of singletons, we find that \east\/ does a better job than
\capthree\/ for error rates larger than $2\%$, while \tgicl\/ does a
better job that \east\/ for error rates larger than $4\%$ to $5\%$.

\begin{figure}[htb]
\begin{minipage}{3in}
\includegraphics[width=3in]{pics.d/numContigs_sanger.pdf}
\end{minipage}
\begin{minipage}{3in}
\includegraphics[width=3in]{pics.d/numSingle_sanger.pdf}
\end{minipage}
\caption{Number of contigs (normalized to the number of transcripts)
  and number of singletons as a function of base-call error
  rate for simulated Sanger sequencing.  \east=blue, \capthree=Black,
  \tgicl=green.  (Number of singletons for
  \velvet\/ and \mira\/ not shown.)}
\label{contigsSanger}
\end{figure}

\noindent {\bf Next Generation Sequences:} In
Figure~\ref{nextgenAscore} we compare the normalized a-score for
simulated 454 and Illumina outputs (with sequence lengths averaging
250 and 62 bp respectively).  For these tests we varied coverage
levels but otherwise continued to use the default Metasim model
\cite{Richter08}.  On 454 data we see significant improvement in
results quality in \east\/ for all levels of coverage.  On Illumina
data we see that \east\/ is competitive with \velvet\/ at higher
coverages, though is unable to match that tool when coverage is
sparse.

\begin{figure}[htb]
\centerline{\includegraphics[width=6in]{pics.d/ascore_nxtgen.pdf}}
\caption{Comparative normalized a-scores for 454 and Illumina sequence
  output, using the \metasim default model at different levels
  of coverage.  \east=blue, \capthree=Black, \tgicl=green.  }
\label{nextgenAscore}
\end{figure}

\noindent {\bf Hybrid:} In Figure~\ref{hybridAscore} we look at the
quality of results derived from applying each tool to a hybrid data
set of Sanger (40\%), 454 (40\%) and Illumina (20\%) sequences, again
showing normalized a-score as a function of base-call error rate in
the Sanger portion.  Here we observe \east\/ matching or outperforming
\capthree\/ and \mira\/ at moderate to high coverages.  At a coverage
level of 50, \east\/ maintains a normalized a-score hovering around
0.85 at all error levels (showing a 54\% improvement over \mira\/ at a
3\% base-call error rate).  \east\/ actually improves its performance
at a coverage level of 30 (hovering around a normalized a-score of
0.95, showing a 142\% improvement over \mira\/ at the 3\% base-call error
rate), an oddity we attribute to the decreased absolute number of NGS
sequences with the smaller coverage.  At a coverage level of 10
\east's normalized a-score is around 0.37 -- performing considerably
worse than \mira\/ at low error rates (e.g. 0.94 v. 0.40 for
error-free bases), but showing a 52\% improvement over \mira\/ at the
3\% error rate mark and in general showing increased robustness to
error.  Remarkably, it is only in this situation that \capthree\/ is
competitive, with that tool achieving a normalized a-score of 0.62
(v. \east's 0.37) for the 3\% base-call error rate, but loosing out to
\east\/ at greater than a 4.5\% base-call error rate -- a situation
likely due to \east's problems with the 62 bp Illumina sequences (see
below).

\begin{figure}[htb]
\centerline{\includegraphics[width=6in]{pics.d/ascore_hybrid.pdf}}
\caption{Comparative normalized a-scores for a hybrid set consisting
  of Sanger, 454 and Illumina sequence output, varying the base-call
  error rate for Sanger sequences and using the MetaSim default model
  for the 454 and Illumina sequences.  Each data set tested was
  composed of 40\% Sanger Sequences, 40\% 454 Sequences, and 20\%
  Illumina sequences.  \east=blue, \capthree=Black, \tgicl=green,
  \velvet=red, \mira=purple.}
\label{hybridAscore}
\end{figure}

\noindent {\bf Tool Runtime:} In Figures~\ref{runtime.fixed} we see
the average runtime for each of the experiments reflected in
Figures~\ref{sangerAscore}-\ref{hybridAscore} for a coverage of 30
(continuing to report the combined runtime of \peace+\east), while in
Figure~\ref{runtime.coverage} we look at the average runtime for each
tool when applied to simulated data sets generated from a single zebra
fish gene at varying levels of coverage.  For Sanger sequences we
see marked improvement of (\peace+)\east\/ over all other tools for
almost all coverage levels: for the smallest set (coverage = 10)
\east\/ has 1.9 speedup over \capthree\/ and a 3.9 speedup over
\tgicl, while at a more moderate size (coverage=50) there are speedups of
4.7 and 8.4.  This returns back down to 3.0 and 5.2 at the largest set
tested (coverage = 100), as can be seen in the bump around a coverage
of 85 in Figure~\ref{runtime.coverage}.  We are unable to
superficially explain this bump, but have noticed the same effect in
data sets derived from other genes as well.

\vspace{3mm}

\east\/ is less competitive on shorter sequence.  In our simulations,
we find a reduced runtime improvement on the simulated 454 sequences
(averaging 250 bp in length), and a significantly reduced runtime on
the simulated Illumina sequences (exactly 62 bp in length).  However,
as technology advances, the ``short-read'' sequencers are producing
increasingly longer reads: 454 sequencing technology can produce
reads exceeding 400 bp, while Illumina sequences can exceed 100 bp
\cite{Eid09,Li10}.  The loss of runtime in our experiments on 454,
Illumina and the hybrid sequences are due to the shorter sequence
(specifically, the problems with applying the $d^2$ metric to shorter
sequences -- as discussed below), hence are becoming less of an issue
with this increase in short-read fragment length.

\vspace{3mm}

In short, we argue that the high quality of our results still makes
\east\/ an appealing option for 454 and hybrid data (while
acknowledging that \east\/ cannot currently compete with \velvet\/ on pure
Illumina data), and also observe that while it has not been
implemented, the basic \east\/ algorithm will easily admit
parallelization -- thus increasing the practical limit on data set
size.

\begin{figure}[htb]
\centerline{\includegraphics[width=6in]{pics.d/runtime_fixedsize_sanger.pdf}}
\caption{Average runtime per trial when when generating data for
  Figures~\ref{sangerAscore}-\ref{hybridAscore}  when coverage = 30.
  Note that the exceptionally high speed of \velvet\/ on Illumina sequences make the
  red (center) bars in that plot essentially indistinguishable.
  (\peace+)\east=blue, \capthree=Black, \tgicl=green,
  \velvet=red, \mira=purple.}
\label{runtime.fixed}
\end{figure}

\begin{figure}[htb]
\centerline{\includegraphics[width=6in]{pics.d/runtime_bycoverage_all.pdf}}
\caption{The average runtime when applying each tool to a simulated
  data set derived from a specific zebrafish transcript of length
  14,625 bp.  As we decrease coverage the number of fragments drops,
  hence illustrating the relationship between tool runtime and
  fragment set size.  Each data point represents the average runtime
  from the application of the tool to at least 30 randomly generated
  simulated sets coming from the transcript at the given level of coverage.
  \peace + \east=blue, \capthree=Black, \tgicl=green,
  \velvet=red, \mira=purple.}
\label{runtime.coverage}
\end{figure}

\noindent{\bf Implementation:} \east\/ has been integrated into the
\peace\/ clustering package \cite{Rao10} and is available for
download from the \peace\/ website ({\bf www.peace-tools.org}).
Users can run the GUI provided by that site, and through it
install both the \peace\/ cluster engine and \east\/ assembly engine
on a local machine or remote server (see Figure~\ref{GUI-fig}).
Initial EST data must be in fasta format, though the tool can print
out the final assembly in fasta, ACE, SAM and BAM formats [NEED TO ADD
CITATION TO THE CONVERSION TOOL].

\begin{figure}[htb]
\centerline{\includegraphics[width=6in]{pics.d/GUI1.pdf}}
\caption{Screen shots of the PEACE GUI.  [COMMENT: Need to improve this.]}
\label{GUI-fig}
\end{figure}

@@

{\textcolor{blue}{In this manuscript we have described the \east\/ algorithm and we have run a
comparison of the \peast\/ tool combination against popular tools
from the literature.}  When compared against competing tools on simulated
data using the normalized a-score and runtime metrics, the \peast\/ combination of
tools produce the best sequence assemblies for Sanger sequence, the
best tool overall [COMMENT: I need a better way to phrase this --
something to indicate the do better as a whole, even if they don't do
as well as some tools on some sets], and by far the most robust against sequencing
error.  While certain tools may do better than \east\/ in specific circumstances,
none of them do better than EAST on multiple sequencing
technology platforms.  While its speed is slower when applied to short
sequence fragments, the quality of its results more then compensate.}
\vspace{3mm}

As observed in the results section, when applying \east, \capthree,
and \tgicl\/ to Sanger sequenced transcripts with normal to high error
rates, \east\/ consistently out performs all other tools in terms of
the normalized a-score metric (Figures~\ref{sangerAscore},
\ref{runtime.fixed}, and \ref{runtime.coverage}).  We also find that
the number of contigs produced is almost exactly equal to the number
of transcripts represented in the set, regardless of error rate, while
the number of singleton ESTs remains low up to about a 5\% error rate
(Figures~\ref{contigsSanger}).  The high normalized a-score indicates
that \east\/ is both reconstructing larger portions of each transcript
and doing so more faithfully, while the low number of contigs means
that we are assembling complete transcripts.  Finally, runtimes for
\east\/ are considerably faster than other tools for larger sets
(Figures~\ref{runtime.fixed} and \ref{runtime.coverage}), with
(\peace+)\east\/ running a minimum of two to three times as fast as
the other tools.  The speedup over \capthree\/ (the faster tool)
ranged from 1.9 to 6.0, depending on the coverage level.  In short,
when used for Sanger Sequences \east\/ is a faster tool that produces
the high quality reconstructions versus comparing tools and is
significantly more robust to error than other tools.

\vspace{3mm}
